Descriptive norms caused increases in mask wearing during the COVID-19 pandemic

### 2023-07-06 manuscript.Rmd
second round of reviews

---
title             : "Descriptive, not injunctive, social norms caused increases in mask wearing during the COVID-19 pandemic"
shorttitle        : "Norms and mask wearing"

@@

abstract: |
 Human sociality is governed by two types of social norms: injunctive norms, which prescribe what people _ought_ to do, and descriptive norms, which reflect what people _actually_ do. The process by which these norms emerge and their causal influences on cooperative behavior over time are not well understood. Here, we study these questions through social norms influencing mask wearing during the COVID-19 pandemic. Leveraging two years of data from the United States (18 time points; _n_ = `r nrow(tar_read(d))`), we tracked mask wearing and perceived injunctive and descriptive mask wearing norms as the pandemic unfolded. Longitudinal trends suggested that norms and behavior were tightly coupled, changing quickly in response to public health recommendations. In addition, longitudinal modelling revealed that descriptive norms, not injunctive norms, caused future increases in mask wearing. During uncertain times, cooperative behavior is driven by what others are actually doing, rather than what others think ought to be done.

note              : "\\footnotesize \\raggedright \\mbox{* indicates shared first authorship, † indicates shared senior authorship} \\par Correspondence concerning this article should be addressed to Samantha L. Heiman, 1101 E 10th St, Bloomington, IN 47405, United States. E-mail: slheiman@iu.edu \\par This study was funded by the Interdisciplinary Cooperation Initiative, ASU President's Office, the Cooperation Science Network, the Institute for Mental Health Research, the University of New Mexico, the Indiana University College of Arts & Sciences, the Rutgers University Center for Human Evolutionary Studies, the Charles Koch Foundation, and the John Templeton Foundation. \\par This working paper has not yet been peer-reviewed."
keywords          : "social norms; descriptive norms; injunctive norms; longitudinal; COVID-19; mask wearing; cooperation"
wordcount         : "4381 words"
bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man, donotrepeattitle"
output            : papaja::apa6_pdf
header-includes: 
  - \usepackage{array}
  - \usepackage{caption}
  - \usepackage{graphicx}
  - \usepackage{siunitx}
  - \usepackage[normalem]{ulem}
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{hhline}
  - \usepackage{calc}
  - \usepackage{tabularx}
  - \usepackage{threeparttable}
  - \usepackage{wrapfig}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \raggedbottom
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1,small}}
  - \captionsetup[figure]{font={stretch=1,small}}
appendix          : "appendix.Rmd"
csl               : nature.csl
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
tar_load(c(fileData, fileOwid, d, owid,
           constructValDes, constructValInj,
           m1.1, m1.2, m2.1, m2.2, m2.3,
           conf1.1, conf1.2, conf2.1, conf2.2, conf2.3,
           pars2.1, pars2.2, pars2.3, riclpm, fitMeasures,
           plot1, plot2, plot3, plot4, plot5, plot6, plot7,
           plot8, itemTable, changePointsTable, lavaanTable))
```

@@

Research has investigated how social norms emerge in a population over time. In the long term, cultural evolutionary models show that injunctive social norms can be vertically transmitted across generations by imitation or teaching, or horizontally diffused from neighboring populations [@Henrich2015]. However, less is known about how social norms arise endogenously within populations in the short term. While researchers have simulated the emergence of descriptive norms [@Muldoon2014a;@Muldoon2014b], this modelling work does not capture how descriptive norms develop alongside injunctive norms in real-world settings. Recent work in behavioral economics has also suggested that injunctive norms of public good provisioning develop in tandem with cooperative behavior through repeated interactions [@Titlestad2019]. But it remains unclear whether these findings generalize beyond the laboratory to real human populations.

With regards to normative influences on behavior, studies have demonstrated positive effects of descriptive norms on a variety of cooperative behaviors, including recycling [@Nigbur2010], paying taxes [@Larkin2018], and sustainably reusing towels in hotels [@Goldstein2008]. However, these studies have two key aspects that limit their ability to assess the causal impact of norms, both of which we address in our current work. First, studies have not accounted for other potential non-social explanations for behavior, such as perceptions of the effectiveness of the behavior and personal beliefs that one should behave in a certain way. These non-social beliefs, labeled "factual beliefs" and "personal normative beliefs" [@Bicchieri2014], often correlate with descriptive and injunctive norms, but they are fundamentally different because they can cause behavior separately from social expectations about what others do or think should be done. For example, willingness to recycle might be driven by perceptions that recycling has a positive impact on the environment and/or personal beliefs that recycling is the right thing to do, even if social norms actively discourage recycling (e.g., recycling is not a common or socially approved behavior). It is thus important to control for non-social beliefs in studies of social norms. Second, studies have tended to follow cross-sectional experimental designs in which social norm perceptions are manipulated by the researchers. However, social norms are not static: they change dynamically over time through processes of deliberation and interaction [@Smith2015]. An alternative but underutilized method of assessing causality between social norms and cooperative behavior, while retaining ecological validity, is to follow these variables over time amidst a real, unfolding social dilemma.

@@

Recent research has found positive relationships between social norms and protective COVID-19 behaviors. In the United States, one study found that perceptions of injunctive norms positively predicted intentions to stay at home to minimize exposure [@Macy2021], and another study found that experimentally-induced descriptive norms increased mask wearing intentions [@Bokemper2021]. In Germany, a two-wave study found that perceptions of descriptive norms positively predicted future protective behaviors, such as physical distancing [@Rudert2021]. These studies are telling, but since they are cross-sectional or only minimally longitudinal, they do not have the temporal granularity to capture fluctuating changes in norm strength and adherence across the pandemic. Furthermore, several of the studies do not control for political ideology, which is important to account for since COVID-19 was highly politicized in the United States [@BaxterKing2022].

Here, we use two years of data from a representative sample of adults in the United States (18 time points; _n_ = 916) to track the development of descriptive and injunctive mask wearing norms and mask wearing behavior over the course of the COVID-19 pandemic. Participants reported their frequency of mask wearing during in-person interactions, as well as their perceptions of descriptive and injunctive mask wearing norms. We also asked participants about their non-social mask wearing beliefs and political ideology, and controlled for these factors. We used these longitudinal data to answer two main research questions in a specific real-world context. First, how do descriptive and injunctive mask wearing norms emerge over time? Second, how do descriptive and injunctive mask wearing norms causally influence mask wearing?

@@

To understand how mask wearing social norms emerged and fluctuated over the course of the COVID-19 pandemic, we first visualized the average descriptive trends of self-reported norm perceptions across the entire study duration. Figure \@ref(fig:plotTimeline) plots self-reported mask wearing and perceptions of descriptive and injunctive mask wearing norms alongside relevant pandemic-related events in the United States, such as CDC public health recommendations and COVID-19 case numbers. These events were obtained from the CDC Museum’s COVID-19 Timeline [@CDC2022].

@@

```{r plotTimeline, echo=F, fig.width=6, fig.height=6, fig.cap="(ref:plotTimelineCaption)"}
plot3
```

@@

Sample averages can provide informative trends, but they do not allow us to estimate within-person changes in mask wearing and perceived social norms over time. To determine whether within-person changes in social norms temporally preceded within-person changes in mask wearing, we fitted a ten-wave unconstrained random-intercept cross-lagged panel model to the longitudinal data. This structural equation model separately estimated stable trait-like between-person individual differences and within-person fluctuations from those trait levels for our main variables (self-reported mask wearing, perceived descriptive mask wearing norms, and perceived injunctive mask wearing norms) and time-variant control variables (factual beliefs and personal normative beliefs). In line with our proposed causal model (Supplementary Figure \@ref(fig:plotDAG)), we also included political orientation as an exogenous time-invariant control. According to established fit statistics, this model fitted the data well (root mean square error of approximation = `r printnum(fitMeasures['rmsea'], digits = 3)`, 95% CI [`r printnum(fitMeasures['rmsea.ci.lower'], digits = 3)` `r printnum(fitMeasures['rmsea.ci.upper'], digits = 3)`]; standardized root mean squared residual = `r printnum(fitMeasures['srmr'], digits = 3)`; comparative fit index = `r printnum(fitMeasures['cfi'], digits = 3)`). Since we are primarily interested in the causal effects of social norms on behavior, in what follows we focus on the results for mask wearing, perceived descriptive norms, and perceived injunctive norms (but see Supplementary Table \@ref(tab:lavaanTable) for full list of estimated autoregressive and cross-lagged effects).

```{r, echo = FALSE}
sc <- standardizedSolution(riclpm)
pe <- parameterEstimates(riclpm)
```

Regarding between-person individual differences, the covariances between the random intercepts in the model revealed positive correlations between stable trait levels of mask wearing and perceived social norms. On average across the whole study, participants who more frequently wore masks during in-person interactions also perceived stronger descriptive mask wearing norms (_r_ = `r printnum(sc[sc$lhs == "ri1" & sc$op == "~~" & sc$rhs == "ri3","est.std"])`, 95% CI [`r printnum(sc[sc$lhs == "ri1" & sc$op == "~~" & sc$rhs == "ri3","ci.lower"])` `r printnum(sc[sc$lhs == "ri1" & sc$op == "~~" & sc$rhs == "ri3","ci.upper"])`], _p_ = `r printp(pe[pe$lhs == "ri1" & pe$op == "~~" & pe$rhs == "ri3","pvalue"])`) and stronger injunctive mask wearing norms (_r_ = `r printnum(sc[sc$lhs == "ri1" & sc$op == "~~" & sc$rhs == "ri2","est.std"])`, 95% CI [`r printnum(sc[sc$lhs == "ri1" & sc$op == "~~" & sc$rhs == "ri2","ci.lower"])` `r printnum(sc[sc$lhs == "ri1" & sc$op == "~~" & sc$rhs == "ri2","ci.upper"])`], _p_ `r printp(pe[pe$lhs == "ri1" & pe$op == "~~" & pe$rhs == "ri2","pvalue"])`). Stable trait perceptions of descriptive and injunctive mask wearing norms were also highly positively correlated (_r_ = `r printnum(sc[sc$lhs == "ri2" & sc$op == "~~" & sc$rhs == "ri3","est.std"])`, 95% CI [`r printnum(sc[sc$lhs == "ri2" & sc$op == "~~" & sc$rhs == "ri3","ci.lower"])` `r printnum(sc[sc$lhs == "ri2" & sc$op == "~~" & sc$rhs == "ri3","ci.upper"])`], _p_ `r printp(pe[pe$lhs == "ri2" & pe$op == "~~" & pe$rhs == "ri3","pvalue"])`).

Regarding within-person dynamics over time, Figure \@ref(fig:plotRICLPM) displays autoregressive and cross-lagged effects for perceived descriptive norms, perceived injunctive norms, and mask wearing across the study duration, controlling for non-social beliefs and political orientation. In random intercept cross-lagged panel models, autoregressive effects represent "persistence" or "inertia" in within-person fluctuations from stable trait levels. In other words, a positive autoregressive effect indicates that being higher than average on one measure predicts being higher than average on that same measure in the following time point (this is not to be confused with the "stable trait level" over time, which is captured by the random intercepts in our model). For example, an autoregressive effect from mask wearing in February 2021 to future mask wearing in June 2021 would suggest that wearing masks more than average in February predicts wearing masks more than average in June. By contrast, and most relevant for the current study, cross-lagged effects represent the effect of a within-person fluctuation in one measure on future within-person fluctuations in other measures. In other words, a positive cross-lagged effect indicates that being higher than average on one measure predicts being higher than average on _another_ measure in the following time point. For example, a cross-lagged effect from descriptive norms in February 2021 to future mask wearing in June 2021 would suggest that perceiving descriptive norms as stronger than average in February predicts wearing masks more than average in June. Cross-lagged effects are thus used to infer within-person causal influences over time.

(ref:plotRICLPMCaption) _Results of ten-wave unconstrained random-intercept cross-lagged panel model._ Circles represent data collection time points. Arrows represent within-person autoregressive effects (on one horizontal level) and cross-lagged effects (across levels) for mask wearing and perceived descriptive and injunctive norms, partitioning out stable between-person individual differences and controlling for factual beliefs, personal normative beliefs, and political orientation. Arrow thickness is scaled according to standardized effect size. Bolded arrows indicate significantly positive parameters, _p_ < 0.05. Gray arrows indicate non-significant parameters. There are no significant direct paths from injunctive norms to future mask wearing, showing that people’s beliefs about what they should be doing did not have any direct influences on future mask wearing. On the other hand, there are significant paths from descriptive norms to future mask wearing, meaning that people’s beliefs about what others are doing influenced their future mask wearing.

```{r plotRICLPM, echo=F, fig.width=7.5, fig.height=3.5, fig.cap="(ref:plotRICLPMCaption)"}
plot7
```

In late 2020 and throughout 2021, we see several cross-lagged effects from perceived descriptive norms to future mask wearing. On four occasions, within-person increases in perceived descriptive norms predicted future within-person increases in mask wearing. According to recent effect size guidelines for cross-lagged panel models [@Orth2022], the standardized beta coefficients for these cross-lagged effects were large (first wave, $\beta$ = `r printnum(sc[sc$lhs == "w1_05" & sc$op == "~" & sc$rhs == "w3_02","est.std"])`, 95% CI [`r printnum(sc[sc$lhs == "w1_05" & sc$op == "~" & sc$rhs == "w3_02","ci.lower"])` `r printnum(sc[sc$lhs == "w1_05" & sc$op == "~" & sc$rhs == "w3_02","ci.upper"])`], _p_ = `r printp(pe[pe$lhs == "w1_05" & pe$op == "~" & pe$rhs == "w3_02","pvalue"])`; second wave, $\beta$ = `r printnum(sc[sc$lhs == "w1_09" & sc$op == "~" & sc$rhs == "w3_05","est.std"])`, 95% CI [`r printnum(sc[sc$lhs == "w1_09" & sc$op == "~" & sc$rhs == "w3_05","ci.lower"])` `r printnum(sc[sc$lhs == "w1_09" & sc$op == "~" & sc$rhs == "w3_05","ci.upper"])`], _p_ = `r printp(pe[pe$lhs == "w1_09" & pe$op == "~" & pe$rhs == "w3_05","pvalue"])`; fourth wave, $\beta$ = `r printnum(sc[sc$lhs == "w1_13" & sc$op == "~" & sc$rhs == "w3_11","est.std"])`, 95% CI [`r printnum(sc[sc$lhs == "w1_13" & sc$op == "~" & sc$rhs == "w3_11","ci.lower"])` `r printnum(sc[sc$lhs == "w1_13" & sc$op == "~" & sc$rhs == "w3_11","ci.upper"])`], _p_ = `r printp(pe[pe$lhs == "w1_13" & pe$op == "~" & pe$rhs == "w3_11","pvalue"])`; fifth wave, $\beta$ = `r printnum(sc[sc$lhs == "w1_14" & sc$op == "~" & sc$rhs == "w3_13","est.std"])`, 95% CI [`r printnum(sc[sc$lhs == "w1_14" & sc$op == "~" & sc$rhs == "w3_13","ci.lower"])` `r printnum(sc[sc$lhs == "w1_14" & sc$op == "~" & sc$rhs == "w3_13","ci.upper"])`], _p_ = `r printp(pe[pe$lhs == "w1_14" & pe$op == "~" & pe$rhs == "w3_13","pvalue"])`). In 2022, these cross-lagged effects from descriptive norms to mask wearing diminished. We also find some evidence for a reciprocal effect, whereby within-person increases in mask wearing predicted future within-person increases in perceived descriptive norms. Moreover, several cross-lagged effects emerged between perceived descriptive and injunctive norms, demonstrating reciprocal within-person causal effects between these variables.

However, the model also reveals that, after controlling for perceived descriptive norms, non-social beliefs, and political orientation, within-person changes in perceived injunctive norms did _not_ predict future within-person changes in mask wearing across our study duration. All cross-lagged effects from perceived injunctive norms to mask wearing are non-significant. Any causal effect that perceived injunctive norms might have had on future mask wearing appears to be fully mediated by perceived descriptive norms. This means that believing that others think that mask wearing is the right thing to do influences one’s perception of what others are actually doing, which then influences future behavior. For example, between August 2021 and December 2021, perceived injunctive norms predicted future perceived descriptive norms, which themselves predicted future mask wearing. But aside from these indirect effects, perceived injunctive norms did not have a direct causal effect on mask wearing over time within individuals.

>>>>

---
title             : "Descriptive norms caused increases in mask wearing during the COVID-19 pandemic"
shorttitle        : "Norms and mask wearing"

@@

abstract: |
 Human sociality is governed by two types of social norms: injunctive norms, which prescribe what people _ought_ to do, and descriptive norms, which reflect what people _actually_ do. The process by which these norms emerge and their causal influences on cooperative behavior over time are not well understood. Here, we study these questions through social norms influencing mask wearing during the COVID-19 pandemic. Leveraging two years of data from the United States (18 time points; _n_ = `r nrow(tar_read(d))`), we tracked mask wearing and perceived injunctive and descriptive mask wearing norms as the pandemic unfolded. Longitudinal trends suggested that norms and behavior were tightly coupled, changing quickly in response to public health recommendations. In addition, longitudinal modeling revealed that descriptive norms caused future increases in mask wearing across multiple waves of data collection. These cross-lagged causal effects of descriptive norms were large, even after controlling for non-social beliefs and demographic variables. Injunctive norms, by contrast, had less frequent and generally weaker causal effects on future mask wearing. During uncertain times, cooperative behavior is more strongly driven by what others are actually doing, rather than what others think ought to be done.

keywords          : "social norms; descriptive norms; injunctive norms; longitudinal; COVID-19; mask wearing; cooperation"
wordcount         : "Main text (excluding Methods): 4553 words"
bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man, donotrepeattitle"
output            : papaja::apa6_pdf
header-includes: 
  - \usepackage{array}
  - \usepackage{caption}
  - \usepackage{graphicx}
  - \usepackage{siunitx}
  - \usepackage[normalem]{ulem}
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{hhline}
  - \usepackage{calc}
  - \usepackage{tabularx}
  - \usepackage{threeparttable}
  - \usepackage{wrapfig}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \raggedbottom
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1,small}}
  - \captionsetup[figure]{font={stretch=1,small}}
  - \note{\raggedright * indicates shared first authorship, † indicates shared senior authorship \par Correspondence concerning this article should be addressed to Samantha L. Heiman, 1101 E 10th St, Bloomington, IN 47405, United States. \text{E-mail:} \href{mailto:slheiman@iu.edu}{\nolinkurl{slheiman@iu.edu}} \par This study was funded by the Interdisciplinary Cooperation Initiative, ASU President's Office, the Cooperation Science Network, the Institute for Mental Health Research, the University of New Mexico, the Indiana University College of Arts and Sciences, the Rutgers University Center for Human Evolutionary Studies, the Charles Koch Foundation, and the John Templeton Foundation.}
appendix          : "appendix.Rmd"
csl               : nature.csl
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
tar_load(
  c(
    d, owid, constructVal, intFreq24hrs, intFreq7days,
    m1.1, m1.2, m2.1, m2.2, m2.3, conf1.1, conf1.2, 
    conf2.1, conf2.2, conf2.3, pars2.1, pars2.2, pars2.3,
    riclpm1_uncon, riclpm2_uncon, riclpm3_uncon,
    riclpm1_con, riclpm2_con, riclpm3_con, riclpm4_con,
    compare1, compare2, compare3, fitMeasures1_uncon, 
    fitMeasures2_uncon, fitMeasures3_uncon,
    fitMeasures1_con, fitMeasures2_con, fitMeasures3_con, 
    fitMeasures4_con, plot01, plot02, plot03, plot04, plot05, 
    plot06, plot07, plot08, plot09, plot10, plot11, plot12, plot13, plot14,
    constrainedTable1, constrainedTable2, constrainedTable3,
    constrainedTable4, itemTable, changePointsTable, unconstrainedTable1,
    unconstrainedTable2, unconstrainedTable3, constructValTable
  )
)
```

@@

Research has investigated how social norms emerge in a population over time. In the long term, cultural evolutionary models show that injunctive social norms can be vertically transmitted across generations by imitation or teaching, or horizontally diffused from neighboring populations [@Henrich2015]. However, less is known about how social norms arise endogenously within populations in the short term. While researchers have simulated the emergence of descriptive norms [@Muldoon2014a;@Muldoon2014b], this modeling work does not capture how descriptive norms develop alongside injunctive norms in real-world settings. Recent work in behavioral economics has also suggested that injunctive norms of public good provisioning develop in tandem with cooperative behavior through repeated interactions [@Titlestad2019]. But it remains unclear whether these findings generalize beyond the laboratory to real human populations.

With regards to normative influences on behavior, studies have demonstrated positive effects of descriptive norms on a variety of cooperative behaviors, including recycling [@Nigbur2010], paying taxes [@Larkin2018], and sustainably reusing towels in hotels [@Goldstein2008]. However, these studies have two key aspects that limit their ability to assess the causal impact of norms, both of which we address in our current work. First, studies have not accounted for other potential non-social explanations for behavior, such as perceptions of the effectiveness of the behavior and personal beliefs that one should behave in a certain way. These non-social beliefs, labeled "factual beliefs" and "personal normative beliefs" [@Bicchieri2014a], often correlate with descriptive and injunctive norms, but they are fundamentally different because they can cause behavior separately from social expectations about what others do or think should be done. For example, willingness to recycle might be driven by perceptions that recycling has a positive impact on the environment and/or personal beliefs that recycling is the right thing to do, even if social norms actively discourage recycling (e.g., recycling is not a common or socially approved behavior). It is thus important to control for non-social beliefs in studies of social norms, especially considering that personal norms have previously been shown to influence prosocial behavior[@Bicchieri2010;@Capraro2021]. Second, studies have tended to follow cross-sectional experimental designs in which social norm perceptions are manipulated by the researchers. However, social norms are not static: they change dynamically over time through processes of deliberation and interaction [@Smith2015]. An alternative but underutilized method of assessing causality between social norms and cooperative behavior, while retaining ecological validity, is to follow these variables over time amidst a real, unfolding social dilemma.

@@

Recent research has studied the relationships between social norms and protective COVID-19 behaviors. In the United States, one study found that perceptions of injunctive norms positively predicted intentions to stay at home to minimize exposure [@Macy2021], and another study found that experimentally-induced descriptive norms increased mask wearing intentions [@Bokemper2021]. In Italy, an experimental study found no effect of messages highlighting descriptive norms, injunctive norms, or personal norms on time spent reading information about COVID-19 governmental rules [@Bilancini2020]. In Germany, a two-wave study found that perceptions of descriptive norms positively predicted future protective behaviors, such as physical distancing [@Rudert2021]. These studies are informative, but since they are cross-sectional or only minimally longitudinal, they do not have the temporal granularity to capture fluctuating changes in norm strength and adherence across the pandemic. Furthermore, several of the studies do not control for potential confounding variables, such as demographics and political ideology. These variables are important to account for as they have previously been shown to be related to COVID-19 attitudes and behaviors [@BaxterKing2022;@Heiman2022].

Here, we use two years of data from a representative sample of adults in the United States (18 time points; _n_ = `r nrow(tar_read(d))`) to track the development of descriptive and injunctive mask wearing norms and mask wearing behavior over the course of the COVID-19 pandemic. Participants reported their frequency of mask wearing during in-person interactions, as well as their perceptions of descriptive and injunctive mask wearing norms. We also asked participants about their non-social mask wearing beliefs, demographics, and political ideology, and controlled for these factors. We used these longitudinal data to answer two main research questions in a specific real-world context. First, how do descriptive and injunctive mask wearing norms emerge over time? Second, how do descriptive and injunctive mask wearing norms causally influence mask wearing?

@@

To answer our first research question about the emergence of mask wearing norms, we first visualized the average descriptive trends of self-reported norm perceptions across the entire study duration, showing how mask wearing social norms emerged and fluctuated over the course of the COVID-19 pandemic. Figure \@ref(fig:plotTimeline) plots self-reported mask wearing and perceptions of descriptive and injunctive mask wearing norms alongside relevant pandemic-related events in the United States, such as CDC public health recommendations and COVID-19 case numbers. These events were obtained from the CDC Museum’s COVID-19 Timeline [@CDC2022].

@@

```{r plotTimeline, echo=F, fig.width=6, fig.height=6, fig.cap="(ref:plotTimelineCaption)"}
plot03
```

@@

Sample averages can provide informative trends, but they do not allow us to determine whether within-person changes in social norms caused future within-person changes in mask wearing over time. To answer our second research question about causal effects, we fitted a ten-wave random-intercept cross-lagged panel model[@Hamaker2015;@Mulder2021] to the longitudinal data. This model separately estimates stable trait-like between-person individual differences and within-person fluctuations from those trait levels for our main variables (self-reported mask wearing, perceived descriptive mask wearing norms, and perceived injunctive mask wearing norms) and control variables (factual beliefs and personal normative beliefs). In line with our proposed causal model (Supplementary Figure \@ref(fig:plotDAG)), we also control for potential confounding in this model by including demographics (gender, age, ethnicity, socioeconomic status) and political orientation as exogenous controls.

Random-intercept cross-lagged panel models capture within-person changes over time with autoregressive and cross-lagged effects. Autoregressive effects represent "persistence" or "inertia" in within-person fluctuations from stable trait levels. In other words, a positive autoregressive effect indicates that being higher than average on one measure predicts being higher than average on that same measure in the following time point (this is not to be confused with the "stable trait level" over time, which is captured by the random intercepts in our model). For example, an autoregressive effect from mask wearing in February 2021 to future mask wearing in June 2021 would suggest that wearing masks more than average in February predicts wearing masks more than average in June. By contrast, and most relevant for the current study, cross-lagged effects represent the effect of a within-person fluctuation in one measure on future within-person fluctuations in other measures. In other words, a positive cross-lagged effect indicates that being higher than average on one measure predicts being higher than average on _another_ measure in the following time point. For example, a cross-lagged effect from descriptive norms in February 2021 to future mask wearing in June 2021 would suggest that perceiving descriptive norms as stronger than average in February predicts wearing masks more than average in June. Cross-lagged effects are thus used to infer within-person causal influences over time. In what follows, we focus on the within-person autoregressive and cross-lagged effects for mask wearing and perceived social norms (see Supplementary Results for between-person results).

```{r echo=FALSE}
# parameter estimates
pe_con <- parameterEstimates(riclpm1_con)
# cross-lagged effects from constrained model
crossInj_con <- pe_con[pe_con$lhs == "w1_05" & pe_con$rhs == "w2_02",]
crossDes_con <- pe_con[pe_con$lhs == "w1_05" & pe_con$rhs == "w3_02",]
```

We first fitted a time-invariant model, which constrained the autoregressive effects, cross-lagged effects, covariances, and variances to equality over time. This model assumes that the relationships between variables are identical across all time points, giving us a sense of the average causal effects of social norms on mask wearing over time. According to established fit statistics, the time-invariant model fitted the data reasonably well (root mean square error of approximation (RMSEA) = `r printnum(fitMeasures1_con['rmsea'], digits = 3)`, 95% CI [`r printnum(fitMeasures1_con['rmsea.ci.lower'], digits = 3)` `r printnum(fitMeasures1_con['rmsea.ci.upper'], digits = 3)`]; standardized root mean squared residual (SRMR) = `r printnum(fitMeasures1_con['srmr'], digits = 3)`; comparative fit index (CFI) = `r printnum(fitMeasures1_con['cfi'], digits = 3)`). Table \@ref(tab:constrainedTable1) summarizes the autoregressive and cross-lagged parameters from the time-invariant model. Autoregressive effects were significantly positive for mask wearing, perceived descriptive norms, and perceived injunctive norms, indicating that being higher than average on these variables at time $t$ generally predicted being higher than average on the same variables at time $t + 1$. Crucially, cross-lagged effects from the time-invariant model revealed that perceived descriptive norms predicted future mask wearing (unstandardized _b_ = `r printnum(crossDes_con[,"est"])`, 95% CI [`r printnum(crossDes_con[,"ci.lower"])` `r printnum(crossDes_con[,"ci.upper"])`], _p_ `r printp(crossDes_con[,"pvalue"])`) while perceived injunctive norms did not (_b_ = `r printnum(crossInj_con[,"est"])`, 95% CI [`r printnum(crossInj_con[,"ci.lower"])` `r printnum(crossInj_con[,"ci.upper"])`], _p_ = `r printp(crossInj_con[,"pvalue"])`). In other words, perceiving descriptive norms as stronger than average at time $t$ generally predicted wearing masks more frequently than average at time $t + 1$, while no such effect existed for injunctive norms. All other cross-lagged effects between mask wearing, perceived descriptive norms, and perceived injunctive norms were significantly positive in the time-invariant model. This general pattern of results was unchanged when removing factual beliefs and personal normative beliefs from the time-invariant model (Supplementary Table \@ref(tab:constrainedTable2)) and when removing factual beliefs, personal normative beliefs, and all exogenous covariates from the time-invariant model (Supplementary Table \@ref(tab:constrainedTable3)).

(ref:rightArrow) →

(ref:constrainedTable1Caption) Unstandardized autoregressive and cross-lagged parameters from time-invariant random-intercept cross-lagged panel model. _Arrows indicate the direction of prediction. Note that the effects of factual beliefs and personal normative beliefs are omitted from this table for clarity. SE = standard error._

```{r constrainedTable1, echo=F, warning=F, message=F}
apa_table(
  constrainedTable1,
  booktabs = TRUE,
  align = c("l", rep("c", 5)),
  caption = "(ref:constrainedTable1Caption)"
)
```

Given that the strength of perceived social norms varied throughout our data collection window (Figure \@ref(fig:plotTimeline)), it is plausible that the causal effects of social norms on mask wearing may have changed over time as well, rather than being identical at each time point. To test whether the equality constraints over time were tenable[@Mulder2021], we compared the time-invariant model to an alternative time-varying model that freely estimated the autoregressive effects, cross-lagged effects, covariances, and variances at each time point. This model assumes that the relationships between variables are different across different time points, allowing the causal effects of social norms on future mask wearing to vary over time. The time-varying model fitted the data better than the time-invariant model (RMSEA = `r printnum(fitMeasures1_uncon['rmsea'], digits = 3)`, 95% CI [`r printnum(fitMeasures1_uncon['rmsea.ci.lower'], digits = 3)` `r printnum(fitMeasures1_uncon['rmsea.ci.upper'], digits = 3)`]; SRMR = `r printnum(fitMeasures1_uncon['srmr'], digits = 3)`; CFI = `r printnum(fitMeasures1_uncon['cfi'], digits = 3)`). Model comparison revealed that freely estimating the parameters over time resulted in improved model fit ($\Delta$AIC = `r printnum(compare1[1,2] - compare1[2,2])`, $\Delta\chi^2$(`r round(compare1[2,6], digits = 0)`) = `r printnum(-compare1[2,5], big.mark = "")`, _p_ `r printp(compare1[2,7])`) suggesting that there was substantial variability in the relationships between mask wearing and social norms across the course of the pandemic. Accordingly, we now turn to the results from the time-varying model.

```{r echo=FALSE}
# parameter estimates and standardized parameters
pe_uncon <- parameterEstimates(riclpm1_uncon)
sc_uncon <- standardizedSolution(riclpm1_uncon)
# unstandardized cross-lagged effects from unconstrained model
crossDes05_unstd_uncon <- pe_uncon[pe_uncon$lhs == "w1_05" & pe_uncon$rhs == "w3_02",]
crossDes09_unstd_uncon <- pe_uncon[pe_uncon$lhs == "w1_09" & pe_uncon$rhs == "w3_05",]
crossDes13_unstd_uncon <- pe_uncon[pe_uncon$lhs == "w1_14" & pe_uncon$rhs == "w3_13",]
crossInj15_unstd_uncon <- pe_uncon[pe_uncon$lhs == "w1_16" & pe_uncon$rhs == "w2_15",]
# standardized cross-lagged effects from unconstrained model
crossDes05_std_uncon <- sc_uncon[sc_uncon$lhs == "w1_05" & sc_uncon$rhs == "w3_02",]
crossDes09_std_uncon <- sc_uncon[sc_uncon$lhs == "w1_09" & sc_uncon$rhs == "w3_05",]
crossDes13_std_uncon <- sc_uncon[sc_uncon$lhs == "w1_14" & sc_uncon$rhs == "w3_13",]
crossInj15_std_uncon <- sc_uncon[sc_uncon$lhs == "w1_16" & sc_uncon$rhs == "w2_15",]
```

Figure \@ref(fig:plotRICLPMPath1) displays the autoregressive and cross-lagged effects for social norms and mask wearing from the time-varying model (see Supplementary Table \@ref(tab:unconstrainedTable1) for full list of estimated autoregressive and cross-lagged effects). In late 2020 and 2021, we find three occasions where within-person increases in perceived descriptive norms predicted future within-person increases in mask wearing. According to recent effect size guidelines for cross-lagged panel models [@Orth2022], the standardized coefficients for these cross-lagged effects were large (first wave, standardized $\beta$ = `r printnum(crossDes05_std_uncon$est.std)`, _b_ = `r printnum(crossDes05_unstd_uncon$est)`, 95% CI [`r printnum(crossDes05_unstd_uncon$ci.lower)` `r printnum(crossDes05_unstd_uncon$ci.upper)`], _p_ = `r printp(crossDes05_unstd_uncon$pvalue)`; second wave, $\beta$ = `r printnum(crossDes09_std_uncon$est.std)`, _b_ = `r printnum(crossDes09_unstd_uncon$est)`, 95% CI [`r printnum(crossDes09_unstd_uncon$ci.lower)` `r printnum(crossDes09_unstd_uncon$ci.upper)`], _p_ = `r printp(crossDes09_unstd_uncon$pvalue)`; fifth wave, $\beta$ = `r printnum(crossDes13_std_uncon$est.std)`, _b_ = `r printnum(crossDes13_unstd_uncon$est)`, 95% CI [`r printnum(crossDes13_unstd_uncon$ci.lower)` `r printnum(crossDes13_unstd_uncon$ci.upper)`], _p_ = `r printp(crossDes13_unstd_uncon$pvalue)`). At other time points, the cross-lagged effects for descriptive norms tended to be estimated in a positive direction, though these estimates were not statistically significant and decreased in magnitude in 2022 (Figure \@ref(fig:plotRICLPMCoef1)). Time-varying models without covariates revealed an additional significant cross-lagged effect of descriptive norms on future mask wearing at the fourth wave, but otherwise the general pattern of results for descriptive norms was unchanged (Supplementary Figures \@ref(fig:plotRICLPMPath2) - \@ref(fig:plotRICLPMCoef3); Supplementary Tables \@ref(tab:unconstrainedTable2) and \@ref(tab:unconstrainedTable3)).
